# Introduction to R and RStudio {#day1}

```{r setup-day1, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Learning Objectives

By the end of Day 1, you will be able to:

- Install and navigate RStudio effectively
- Understand basic R data structures (vectors, data frames, lists)
- Import and explore simple datasets
- Write basic control flow structures and functions

## Module 1: Setting Up and Getting Started with R {#day1-mod1}

### Introduction

R is a powerful programming language and environment specifically designed for statistical computing and graphics. RStudio is an integrated development environment (IDE) that makes working with R much easier.

### Installing R and RStudio

1. **Install R** (version 4.3.0 or higher)
   - Visit [https://cran.r-project.org/](https://cran.r-project.org/)
   - Download the version appropriate for your operating system
   - Run the installer

2. **Install RStudio Desktop**
   - Visit [https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/)
   - Download the free Desktop version
   - Run the installer
   
3. **Install Required R Packages**

    Open an R session in the terminal or RStudio:
    
    ```r
    # Install CRAN packages
    install.packages(c(
      "bookdown", "rmarkdown", "knitr", "pheatmap", "ggplot2", "downlit", "xml2",
      "reshape2", "gridExtra", "tidyverse", "lme4",
      "ggforce", "scatterpie", "png"  # Optional
    ))
    
    # Install Bioconductor packages
    if (!requireNamespace("BiocManager", quietly = TRUE))
      install.packages("BiocManager")
    
    BiocManager::install(c(
      "limma", "vsn", "sva", "clusterProfiler", "org.Hs.eg.db", "lme4",
      "KEGGREST", "AnnotationDbi", "annotate", "GO.db",
      "genefilter", "GOSemSim", "DOSE", "enrichplot"
    ), update = TRUE, ask = FALSE)
    ```

### RStudio Interface Tour

RStudio has four main panes:

1. **Source Editor** (top-left): Where you write and edit your scripts
2. **Console** (bottom-left): Where code is executed and results appear
3. **Environment/History** (top-right): Shows objects in memory and command history
4. **Files/Plots/Packages/Help** (bottom-right): File browser, plot viewer, package manager, and help documentation

### Scripts vs Console

The **Console** is for:
- Quick calculations
- Testing commands
- Interactive exploration

**Scripts** (.R or .Rmd files) are for:
- Saving your work
- Creating reproducible analyses
- Organizing complex workflows

### Basic Operators

```{r basic-operators}
# Arithmetic operators
5 + 3        # Addition
10 - 4       # Subtraction
6 * 7        # Multiplication
20 / 4       # Division
2 ^ 3        # Exponentiation
17 %% 5      # Modulo (remainder)

# Comparison operators
5 == 5       # Equal to
5 != 3       # Not equal to
7 > 3        # Greater than
4 < 8        # Less than
5 >= 5       # Greater than or equal
3 <= 10      # Less than or equal

# Logical operators
TRUE & FALSE  # AND
TRUE | FALSE  # OR
!TRUE         # NOT

# Scientific notation
1.5e6  # 1,500,000
```

**Tip:** Use descriptive variable names with underscores

### Variables & Assignment

```{r}
# Assignment operator
x <- 10      # Assign 10 to x
y = 5        # Alternative (but <- is preferred)

## Assign values to variables
protein_count <- 800
sample_size <- 12
study_name <- "Proteomics_2025"

## View variables
protein_count
sample_size

## Use variables in calculations
total_measurements <- protein_count * sample_size
total_measurements
```

### Data Types

```{r data_types}
# Numeric
intensity <- 25114306.44
class(intensity)

# Character (text)
accession <- "F1LMU0"
class(accession)

# Logical
is_significant <- TRUE
class(is_significant)

# Check type
is.numeric(intensity)
is.character(accession)
```

### Creating Your First Script

```{r first-script, eval=FALSE}
# Create a new R script: File > New File > R Script
# Or use Ctrl+Shift+N (Windows/Linux) or Cmd+Shift+N (Mac)

# Write your code
message("Hello, Proteomics World!")

# Save your script: File > Save
# Run code: Ctrl+Enter (Windows/Linux) or Cmd+Return (Mac)
```

### Exercise 1.1: First Steps

Create a new R script and:

1. Calculate the sum of 123 and 456
2. Assign the result to a variable called `total`
3. Print the value of `total`
4. Calculate what percentage 123 is of the total

```{r exercise-1-1-solution, class.source="fold-hide"}
# Solution
result1 <- 123 + 456
total <- result1
print(total)

percentage <- (123 / total) * 100
print(paste0("123 is ", round(percentage, 2), "% of the total"))
```

## Module 2: Data Types and Structures {#day1-mod2}

### Vectors

Vectors are the most basic data structure in R. They contain elements of the same type.

```{r vectors}
# Numeric vectors
ages <- c(25, 30, 35, 40, 45)
print(ages)

# Character vectors
names <- c("Alice", "Bob", "Charlie", "Diana", "Eve")
print(names)

# Logical vectors
passed_qc <- c(TRUE, TRUE, FALSE, TRUE, TRUE)
print(passed_qc)

# Sequences
seq_1_10 <- 1:10
seq_custom <- seq(from = 0, to = 100, by = 10)

# Vector operations
mean(ages)
```

### Indexing and Subsetting

```{r indexing}
# Access elements by position (1-indexed!)
ages[1]           # First element
ages[c(1, 3, 5)]  # Multiple elements
ages[-2]          # All except second element

# Logical indexing
ages[ages > 35]   # Elements greater than 35

# Named vectors
protein_abundance <- c(ACTB = 1500, GAPDH = 2000, MYC = 800)
protein_abundance["ACTB"]
```

### Data Frames

Data frames are the most common structure for storing tabular data.

```{r dataframes}
# Create a data frame
patient_data <- data.frame(
  patient_id = 1:5,
  name = c("Alice", "Bob", "Charlie", "Diana", "Eve"),
  age = c(25, 30, 35, 40, 45),
  treatment = c("A", "B", "A", "B", "A"),
  response = c(TRUE, TRUE, FALSE, TRUE, FALSE),
  stringsAsFactors = FALSE
)

print(patient_data)

# View structure
str(patient_data)

# Summary statistics
summary(patient_data)

# Access columns
patient_data$age
patient_data[, "name"]
patient_data[, 2]

# Access rows
patient_data[1, ]           # First row
patient_data[1:3, ]         # First three rows

# Access specific cells
patient_data[2, 3]          # Row 2, Column 3
patient_data[2, "age"]      # Same, using column name

# Subset by condition
patient_data[patient_data$age > 30, ]
patient_data[patient_data$treatment == "A", ]
```

#### Exploring Data Frames

```{r explore_df}
# View first rows
head(patient_data, 2)

# Dimensions
dim(patient_data)
nrow(patient_data)
ncol(patient_data)

# Column names
colnames(patient_data)
```

#### Accessing Data Frame Elements

```{r access_df}
# Access column by name
patient_data$name

# Access using brackets [row, column]
patient_data[1, 3]        # Row 1, column 3
patient_data[1:2, ]       # First 2 rows, all columns
patient_data[, "treatment"]    # All rows, Mass column

# Subset data
patient_data[patient_data$age > 30, ]
```

### Lists

Lists can contain elements of different types and structures.

```{r lists}
# Create a list
experiment <- list(
  experiment_id = "EXP001",
  date = "2025-01-15",
  samples = c("S1", "S2", "S3"),
  data = patient_data,
  validated = TRUE
)

# Access list elements
experiment$experiment_id
experiment[[1]]
experiment[["samples"]]
```

### Factors

Factors are used for categorical data.

```{r factors}
# Create factor
treatment_factor <- factor(c("Control", "Drug A", "Drug B", "Control", "Drug A"))
print(treatment_factor)

# Check levels
levels(treatment_factor)

# Ordered factors
severity <- factor(
  c("Mild", "Severe", "Moderate", "Mild", "Severe"),
  levels = c("Mild", "Moderate", "Severe"),
  ordered = TRUE
)
print(severity)
```

### Type Coercion

```{r coercion}
# Implicit coercion
mixed <- c(1, 2, "three", 4)  # All converted to character
print(mixed)

# Explicit coercion
numbers_char <- c("1", "2", "3", "4")
numbers_num <- as.numeric(numbers_char)
print(numbers_num)

# Check types
class(mixed)
is.numeric(mixed)
is.character(mixed)
```

### Exercise 1.2: Data Structures

Create a data frame for a proteomic experiment with:

- 10 protein IDs (P001 to P010)
- Random abundance values between 100 and 5000
- Random p-values between 0 and 1
- Significance status (TRUE if p-value < 0.05)

```{r exercise-1-2-solution, class.source="fold-hide"}
# Solution
set.seed(42)  # For reproducibility

proteins <- data.frame(
  protein_id = paste0("P", sprintf("%03d", 1:10)),
  abundance = round(runif(10, min = 100, max = 5000), 2),
  p_value = runif(10, min = 0, max = 1),
  stringsAsFactors = FALSE
)

proteins$significant <- proteins$p_value < 0.05

print(proteins)

# Summary
cat("\nNumber of significant proteins:", sum(proteins$significant), "\n")
```

## Module 3: Control Flow and Functions {#day1-mod3}

### Conditional Statements

```{r conditionals}
# if statement
x <- 10

if (x > 5) {
  print("x is greater than 5")
}

# if-else
if (x > 15) {
  print("x is greater than 15")
} else {
  print("x is 15 or less")
}

# if-else if-else
score <- 75

if (score >= 90) {
  grade <- "A"
} else if (score >= 80) {
  grade <- "B"
} else if (score >= 70) {
  grade <- "C"
} else {
  grade <- "F"
}

print(paste("Grade:", grade))

# Vectorized ifelse
values <- c(1, 5, 10, 15, 20)
categories <- ifelse(values > 10, "High", "Low")
print(categories)
```

### Loops

```{r loops}
# for loop
for (i in 1:5) {
  print(paste("Iteration:", i))
}

# Loop through vector
proteins <- c("ACTB", "GAPDH", "MYC")
for (protein in proteins) {
  print(paste("Processing:", protein))
}

# while loop
counter <- 1
while (counter <= 5) {
  print(paste("Counter:", counter))
  counter <- counter + 1
}

# Loop with condition
numbers <- 1:10
for (num in numbers) {
  if (num %% 2 == 0) {
    print(paste(num, "is even"))
  }
}
```

### Functions

```{r functions}
# Basic function
greet <- function(name) {
  message <- paste("Hello,", name, "!")
  return(message)
}

greet("Alice")

# Function with multiple parameters
calculate_fold_change <- function(treatment, control) {
  fc <- treatment / control
  log2_fc <- log2(fc)
  return(log2_fc)
}

calculate_fold_change(treatment = 200, control = 100)

# Function with default parameters
normalize_abundance <- function(abundance, method = "median") {
  if (method == "median") {
    normalized <- abundance / median(abundance, na.rm = TRUE)
  } else if (method == "mean") {
    normalized <- abundance / mean(abundance, na.rm = TRUE)
  } else {
    stop("Method must be 'median' or 'mean'")
  }
  return(normalized)
}

values <- c(100, 200, 300, 400, 500)
normalize_abundance(values)
normalize_abundance(values, method = "mean")
```

### Apply Family Functions

```{r apply-functions}
# Create sample data
protein_matrix <- matrix(
  c(100, 150, 200, 250, 
    110, 160, 210, 260,
    120, 170, 220, 270),
  nrow = 3, byrow = TRUE
)
colnames(protein_matrix) <- c("Sample1", "Sample2", "Sample3", "Sample4")
rownames(protein_matrix) <- c("Protein1", "Protein2", "Protein3")

print(protein_matrix)

# apply: apply function to rows or columns
row_means <- apply(protein_matrix, 1, mean)  # 1 = rows
col_means <- apply(protein_matrix, 2, mean)  # 2 = columns

print(row_means)
print(col_means)

# lapply: apply function to list, returns list
my_list <- list(a = 1:5, b = 6:10, c = 11:15)
list_means <- lapply(my_list, mean)
print(list_means)

# sapply: simplified version of lapply
vector_means <- sapply(my_list, mean)
print(vector_means)
```

### Exercise 1.3: Functions and Loops

Write a function that:

1. Takes a vector of protein abundances
2. Calculates the coefficient of variation (CV = sd/mean * 100)
3. Returns "Pass" if CV < 20%, "Fail" otherwise

Apply this function to multiple samples using a loop.

```{r exercise-1-3-solution, class.source="fold-hide"}
# Solution
calculate_cv_status <- function(abundances) {
  cv <- (sd(abundances, na.rm = TRUE) / mean(abundances, na.rm = TRUE)) * 100
  
  if (cv < 20) {
    status <- "Pass"
  } else {
    status <- "Fail"
  }
  
  return(list(cv = round(cv, 2), status = status))
}

# Create sample data
sample_data <- list(
  sample1 = c(100, 105, 98, 102, 99),
  sample2 = c(100, 150, 90, 200, 80),
  sample3 = c(500, 505, 498, 502, 496)
)

# Apply function
for (sample_name in names(sample_data)) {
  result <- calculate_cv_status(sample_data[[sample_name]])
  cat(sample_name, "- CV:", result$cv, "% - Status:", result$status, "\n")
}
```

## Module 4: Importing and Exploring Data

### Reading CSV Files

```{r import-data, eval=FALSE}
# Read CSV
data <- read.csv("data/proteins.csv")

# Read with tidyverse
library(readr)
data <- read_csv("data/proteins.csv")

# Read tab-delimited
data <- read.delim("data/proteins.txt", sep = "\t")
```

### Basic Data Exploration

```{r explore-data}
# Create example data
set.seed(123)
protein_data <- data.frame(
  protein_id = paste0("P", 1:100),
  abundance = rnorm(100, mean = 1000, sd = 200),
  condition = rep(c("Control", "Treatment"), each = 50)
)

# Dimensions
dim(protein_data)
nrow(protein_data)
ncol(protein_data)

# First and last rows
head(protein_data)
tail(protein_data)

# Summary statistics
summary(protein_data)

# Table for categorical data
table(protein_data$condition)
```


## Module 5: Data Wrangling with tidyverse

### Introduction to tidyverse

**What is tidyverse?**

- Collection of R packages for data science
- Consistent syntax and design philosophy
- Main packages: dplyr, tidyr, ggplot2, readr

**Why tidyverse?**

- Readable, intuitive code
- Pipe operator `%>%` for chaining operations
- Faster learning curve
- Industry standard

```{r, setup, include=FALSE}
#knitr::opts_knit$set(root.dir = '/08_pkgs/r4proteomics/') # may need to avoid re-setup
```

```{r load_tidyverse}
library(tidyverse)
```

---

### The Pipe Operator %>%

**Traditional approach:**
```{r no_pipe}
x <- c(1, 2, 3, 4, 5)
result <- mean(log10(sqrt(x)))
result
```

**With pipe operator:**
```{r with_pipe}
result <- x %>% 
  sqrt() %>% 
  log10() %>% 
  mean()
result
```

**Read as:** "Take x, THEN take square root, THEN take log10, THEN calculate mean"

---

### dplyr: Grammar of Data Manipulation

**Key Functions:**

- `select()` - Choose columns
- `filter()` - Choose rows based on conditions
- `mutate()` - Create or modify columns
- `arrange()` - Sort data
- `summarize()` - Calculate summary statistics
- `group_by()` - Group data for operations

---

### Your Actual Proteomics Data Structure

```{r create_sample_data}
# Simulate your data structure
set.seed(123)
proteomics_data <- readxl::read_excel(
    path = "examples/2126001_Protein Export_ex.xlsx", #this should the file example that genethon provided
    skip = 1
)
proteomics_data <- proteomics_data %>%
    select(where(~ any(!is.na(.))))

head(proteomics_data[, 1:10])
```

Good practice:
Convert all column names to lowercase and replace spaces with underscores `_`,
to keep the naming of colmuns cleaner and more consistent.

```{r}
#colnames(proteomics_data) <- tolower(colnames(proteomics_data))
colnames(proteomics_data) <- gsub(" ", "_", tolower(colnames(proteomics_data)))
```


---

### select(): Choose Columns

```{r select_demo}
# Select specific columns
proteomics_data %>%
  select(accession, gene, mass, `2126001_029_f5`) %>%
  head(3)

# Select range of columns
proteomics_data %>%
  select(accession:mass) %>%
  head(3)

# Select columns containing "sample"
proteomics_data %>%
  select(accession, contains("f5")) %>%
  head(3)
```

---

### filter(): Choose Rows

```{r filter_demo}
# Filter high mass proteins
proteomics_data %>%
  filter(mass > 100000) %>%
  select(accession, gene, mass)

# Multiple conditions (AND)
proteomics_data %>%
  filter(mass > 100000 & peptide_count > 100) %>%
  select(accession, gene, mass, peptide_count)

# OR conditions
proteomics_data %>%
  filter(mass > 200000 | peptide_count > 300) %>%
  select(accession, gene, mass, peptide_count)
```

---

### mutate(): Create New Columns

```{r mutate_demo}
# Calculate log10 transformed values
proteomics_data %>%
  mutate(
    log10_mass = log10(mass)
  ) %>%
  select(accession, mass, log10_mass, `2126001_029_f5`) %>%
  head(3)
```

---

### arrange(): Sort Data

```{r arrange_demo}
# Sort by mass (ascending)
proteomics_data %>%
  arrange(mass) %>%
  select(accession, gene, mass) %>%
  head(3)

# Sort descending
proteomics_data %>%
  arrange(desc(mass)) %>%
  select(accession, gene, mass) %>%
  head(3)

# Sort by multiple columns
proteomics_data %>%
  arrange(desc(peptide_count), mass) %>%
  select(accession, peptide_count, mass) %>%
  head(3)
```

---

### summarize(): Calculate Statistics

```{r summarize_demo}
# Calculate summary statistics
proteomics_data %>%
  summarize(
    n_proteins = n(),
    mean_mass = mean(mass),
    median_mass = median(mass),
    sd_mass = sd(mass),
    mean_intensity = mean(`2126001_029_f5`)
  )

# Multiple samples
proteomics_data %>%
  summarize(
    mean_s1 = mean(`2126001_029_f5`),
    mean_s2 = mean(`2126001_359_f9`),
    cv_s1 = sd(`2126001_029_f5`) / mean(`2126001_029_f5`) * 100
  )
```

---

### group_by(): Grouped Operations

```{r groupby_demo}
# Create groups based on mass
proteomics_data_grouped <- proteomics_data %>%
  mutate(
    mass_category = case_when(
      mass < 50000 ~ "Low",
      mass < 150000 ~ "Medium",
      TRUE ~ "High"
    )
  )

# Calculate statistics by group
proteomics_data_grouped %>%
  group_by(mass_category) %>%
  summarize(
    n_proteins = n(),
    mean_peptides = mean(peptide_count),
    mean_confidence = mean(confidence_score)
  )
```

---

### Combining Operations: Pipeline

```{r pipeline_demo}
# Complete analysis pipeline
results <- proteomics_data %>%
  # Filter high confidence proteins
  filter(confidence_score > 10000) %>%
  # Calculate mean intensity across samples
  mutate(
    mean_intensity = rowMeans(select(., starts_with("sample")))
  ) %>%
  # Keep relevant columns
  select(accession, gene, mass, confidence_score, mean_intensity) %>%
  # Sort by mean intensity
  arrange(desc(mean_intensity))

head(results)
```

---

### tidyr: Reshaping Data

**Key Functions:**

- `pivot_longer()` - Wide to long format
- `pivot_wider()` - Long to wide format
- `separate()` - Split one column into multiple
- `unite()` - Combine columns

**Why reshape?** Different analyses and visualizations require different formats

---

### Wide vs Long Format

**Wide format (your current data):**
```
accession  `2126001_029_f5`  `2126001_359_f9`  sample_3
F1LMU0     2511430   8316460   3577492
G3V8V3     1851936   2066635   2986710
```

**Long format:**
```
accession  Sample     Intensity
F1LMU0     `2126001_029_f5`   2511430
F1LMU0     `2126001_359_f9`   8316460
F1LMU0     sample_3   3577492
G3V8V3     `2126001_029_f5`   1851936
```

---

### pivot_longer(): Wide to Long

```{r pivot_longer_demo}
# Convert to long format
proteomics_long <- proteomics_data %>%
  pivot_longer(
    cols = starts_with("2"),
    names_to = "Sample",
    values_to = "Intensity"
  )

head(proteomics_long, 10)
```

---

### Why Long Format?

**Advantages for analysis:**

- Easier grouping and summarizing
- Better for ggplot2 visualizations
- Facilitates statistical modeling
- Standard format for many tools

```{r long_analysis}
# Calculate statistics by sample
proteomics_long %>%
  group_by(Sample) %>%
  summarize(
    mean_intensity = mean(Intensity),
    median_intensity = median(Intensity),
    n_proteins = n()
  ) %>%
  head(6)
```

---

### pivot_wider(): Long to Wide

```{r pivot_wider_demo}
# Convert back to wide format
proteomics_wide <- proteomics_long %>%
  pivot_wider(
    names_from = Sample,
    values_from = Intensity
  )

head(proteomics_wide[, 1:10])
```

---

### Working with Strings

```{r string_demo}
# Extract gene names
proteomics_data %>%
  mutate(
    # Convert to uppercase
    gene_upper = toupper(gene),
    # Detect pattern
    is_kinase = str_detect(description, "kinase"),
    # Extract words
    first_word = word(description, 1)
  ) %>%
  select(gene, description, is_kinase, first_word) %>%
  head(3)
```

---

### Handling Missing Values

```{r missing_values}
# Create data with missing values
data_with_na <- proteomics_data %>%
  mutate(
    `2126001_029_f5` = ifelse(`2126001_029_f5` < quantile(`2126001_029_f5`, 0.2), 
                     NA, `2126001_029_f5`)
  )

# Check missing values
data_with_na %>%
  summarize(
    n_total = n(),
    n_missing_s1 = sum(is.na(`2126001_029_f5`)),
    percent_missing = mean(is.na(`2126001_029_f5`)) * 100
  )

# Remove rows with NA
data_with_na %>%
  filter(!is.na(`2126001_029_f5`)) %>%
  nrow()

# Replace NA with value
data_with_na %>%
  mutate(`2126001_029_f5` = replace_na(`2126001_029_f5`, 0)) %>%
  head(3)
```

---

### Hands-On Exercise

**Task:** Transform and analyze your proteomics data

```{r day2_exercise, eval=FALSE}
# 1. Load your data
my_data <- read_excel("your_data.xlsx")

# 2. Filter and calculate
high_quality <- my_data %>%
  filter(confidence_score > 10000, peptide_count > 50) %>%
  mutate(mean_intensity = rowMeans(select(., starts_with("sample"))))

# 3. Convert to long format
long_data <- my_data %>%
  pivot_longer(cols = starts_with("sample"),
               names_to = "Sample",
               values_to = "Intensity")

# 4. Calculate CV per protein
cv_data <- long_data %>%
  group_by(accession) %>%
  summarize(mean_int = mean(Intensity, na.rm = TRUE),
            sd_int = sd(Intensity, na.rm = TRUE),
            cv = (sd_int / mean_int) * 100)
```

## Day 1 Summary

Today you learned:

- ✓ How to set up R and RStudio
- ✓ Basic R operators and syntax
- ✓ Data structures: vectors, data frames, lists, factors
- ✓ Indexing and subsetting data
- ✓ Control flow: if/else, loops
- ✓ Writing custom functions
- ✓ Importing and exploring data

### Homework

1. Install all required packages for Day 2
2. Practice writing functions for data manipulation
3. Explore the built-in datasets in R (use `data()` to see available datasets)

```{r homework-packages, eval=FALSE}
# Install packages for Day 2
install.packages(c("ggplot2", "dplyr", "tidyr", "pheatmap"))

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c("limma", "vsn"))
```

## Additional Resources

- [R for Data Science](https://r4ds.had.co.nz/) by Hadley Wickham
- [RStudio Cheat Sheets](https://posit.co/resources/cheatsheets/)
- [Stack Overflow](https://stackoverflow.com/questions/tagged/r) for questions

