# Functional Analysis, Longitudinal & Public Data {#day4}

## Objectives

-   Interpret DE results biologically\
-   Explore longitudinal trajectories\
-   Work with public datasets

## Module 1: Functional Enrichment & GSEA

Introduce enrichment (GO, KEGG, Reactome).\
GSEA principles and visualization.

```{r warning=FALSE, message=FALSE, eval=FALSE}
# example with clusterProfiler
library(clusterProfiler)
library(org.Hs.eg.db)
# Suppose `de_genes` is a vector of gene IDs with statistics
ego <- enrichGO(de_genes, OrgDb = org.Hs.eg.db, keyType = "ENSEMBL", ont = "BP")
dotplot(ego)
```

```{r warning=FALSE}
library(lme4)
df_long <- data.frame(
  sample = rep(1:10, each = 3),
  time = rep(c(0,1,2), times = 10),
  intensity = rnorm(30)
)
m <- lmer(intensity ~ time + (1 + time | sample), data = df_long)
summary(m)
```

### Exercise

On a longitudinal proteomics dataset, plot trajectories per protein or per cluster, fit simple linear/mixed model.

## Module 2: Public Data Integration

Demonstrate downloading, cleaning, integrating a public proteomics or expression dataset (e.g. from PRIDE, GEO). Run pipeline: QC → normalization → DE → enrichment.

### Exercise

Pick a public dataset, import into R, preprocess, analyze, interpret results.
